---
title: "homework1"
author: "Sherry Kawing Lau (GTID: klau37)"
date: "August 27, 2016"
output: pdf_document
---
```{r setup, include=FALSE}
library(knitr)
```

### Question 2

Implement a function that computes the log of the factorial value of an integer using a for loop. Note that implementing it using log(A)+log(B)+ · · · avoids overflow while implementing it as log(A · B · · · · ) creates an overflow early on.

```{r}
logFactorialForLoop = function(n){
  if (n<=0){
    print("Error: Out of Range")
  } else{
    # log(n!) = log(1) + log(2) + log(3) + … + log(n)
    sum_log = 0
    for(i in 1:n){
      sum_log = sum_log + log(i)
    }
    return(sum_log)
  }
}

```

### Question 3

Implement a function that computes the log of the factorial value of an integer using recursion.

```{r}
logFactorialRecursion = function(n){
  if (n<=0){
    print("Error: Out of Range")
  } else if (n==1){
    return(log(1))
  } else{
    sum_log = log(n) + sum(log(seq(n-1)))
    return(sum_log)
  }
}
```

### Question 4

Using your two implementations of log-factorial in (2) and (3) above, compute the sum of the log-factorials of the integers 1, 2, . . . , N for various N values.

```{r}
sumLFForLoop = function(N){
  sumLFForLoop = logFactorialForLoop(N) + sum(unlist(lapply(seq(N-1), logFactorialForLoop)))
  return(sumLFForLoop)
}

sumLFRecursion = function(N){
  sumLFRecursion = logFactorialRecursion(N) + sum(unlist(lapply(seq(N-1), logFactorialRecursion)))
  return(sumLFRecursion)
}

sumLFForLoop(100) # Sum of Log-Factorials for N=100 using For Loop
sumLFRecursion(100) # Sum of Log-Factorials for N=100 using Recursions

```

### Question 5

Compare the execution times of your two implementations for (4) with an implementation based on the official R function lfactorial(n). You may use the function system.time() to measure execution time. What are the growth rates of the three implementations as N increases? Use the command options(expressions=500000) to increase the number of nested recursions allowed. Compare the timing of the recursion implementation as much as possible, and continue beyond that for the other two implementations.

#### Computational Time Comparison for Log Factorial
Comparing the calculation time for log factorial using the three methods: 1. log(n!) with for loop (2), 2. log(n!) with recursion (3), log(n!) with lfactorial function:

```{r}
integer_value = c(1000, 10000, 100000, 500000)

time = NULL
for(i in 1:length(integer_value)){
  LFForLoopTime = system.time(logFactorialForLoop(integer_value[i]))
  LFForLoopTimeRow = data.frame(method="For Loop", 
                                integer=integer_value[i], 
                                logfactorial=logFactorialForLoop(integer_value[i]), 
                                user=LFForLoopTime[1], 
                                system= LFForLoopTime[2], 
                                elapsed=LFForLoopTime[3])
  
  LFRecursionTime = system.time(logFactorialRecursion(integer_value[i]))
  LFRecursionTimeRow = data.frame(method="Recursion", 
                                  integer=integer_value[i], 
                                  logfactorial=logFactorialRecursion(integer_value[i]), 
                                  user=LFRecursionTime[1], 
                                  system= LFRecursionTime[2], 
                                  elapsed=LFRecursionTime[3])
  
  LFTime = system.time(lfactorial(integer_value[i]))
  LFTimeRow = data.frame(method="lfactorial", 
                         integer=integer_value[i], 
                         logfactorial=lfactorial(integer_value[i]), 
                         user=LFTime[1], 
                         system=LFTime[2], 
                         elapsed=LFTime[3])
  
  time[[i]] = do.call(rbind, list(LFForLoopTimeRow, LFRecursionTimeRow, LFTimeRow))
}
time_table = do.call(rbind, time)
row.names(time_table) = NULL
time_table = time_table[order(time_table$method),]
kable(time_table)
```


With for loop, increasing the values of integer increases computation time drastically. With recursion, computation time increases as integer values increase but the growth rate is less than for loop method. For lfactorial function, computation time doesnt change much with 3 decimal places as integer values increases.

#### Computational Time Comparison for Sum of Log Factorial

Performing similar analysis for calculating sum of log factorial: 1. sum(log(n!)) for N with for loop, 2. sum(log(n!)) for N with recursion, 3. sum(log(n!)) for N with lfactorial function

```{r}
sumLF = function(N){
  sumLF = logFactorialRecursion(N) + sum(unlist(lapply(seq(N-1), lfactorial)))
  return(sumLF)
}

integer_value = c(10, 100, 500, 1000)

sum_time = NULL
for(i in 1:length(integer_value)){
  print(i)
  LFForLoopTime = system.time(sumLFForLoop(integer_value[i]))
  LFForLoopTimeRow = data.frame(method="For Loop", 
                                integer=integer_value[i], 
                                logfactorial=sumLFForLoop(integer_value[i]), 
                                user=LFForLoopTime[1], 
                                system= LFForLoopTime[2], 
                                elapsed=LFForLoopTime[3])
  
  LFRecursionTime = system.time(sumLFRecursion(integer_value[i]))
  LFRecursionTimeRow = data.frame(method="Recursion", 
                                  integer=integer_value[i], 
                                  logfactorial=sumLFRecursion(integer_value[i]), 
                                  user=LFRecursionTime[1], 
                                  system= LFRecursionTime[2], 
                                  elapsed=LFRecursionTime[3])
  
  LFTime = system.time(sumLF(integer_value[i]))
  LFTimeRow = data.frame(method="lfactorial", 
                         integer=integer_value[i], 
                         logfactorial=sumLF(integer_value[i]), 
                         user=LFTime[1], 
                         system=LFTime[2], 
                         elapsed=LFTime[3])
  
  sum_time[[i]] = do.call(rbind, list(LFForLoopTimeRow, LFRecursionTimeRow, LFTimeRow))
}
sum_time_table = do.call(rbind, sum_time)
row.names(sum_time_table) = NULL
sum_time_table = sum_time_table[order(sum_time_table$method),]
kable(sum_time_table)


```

Similar to calucating log factorial, for loop shows the highest growth rate in computational time as integer increases, then recursion and lfactorial is the fastest.
